Efficient pedigree recording for fast population genetics simulation

### 2018-01-17 forwards_paper.tex
note on harmonic series
missing 1/k
Fixed parathesis in grant ACK.

%% Progressive memory reduction as a function of T
What if $T$ is smaller:
how many of the resulting $2NT$ edges are required after simplification?
In other words, how fast does the information in the pedigree become irrelevant?
% \citet{padhukasahasram2008exploring} obtained impressive speedups relative to contemporary simulators
% by retaining only 8 generations of history.
% This begs the question: how fast does the information in the pedigree become irrelevant?
Now, we need to compute the expected total length of all branches in a coalescent tree
up until time $T$ (or the common ancestor, whichever comes first).
Again, coalescent theory tells us that
the expected length of time for which a coalescent tree has $k$ lineages
is $2N/(k(k-1)) = 2N(1/(k-1) - 1/k)$ generations.
Since the tree has $k$ branches over this period,
it is expected to contribute $2N/(k-1)$ to the total tree length.
By summing over $n < k \le N$,
the $N$ tips of a tree are expected to descend from only $n$ lineages
around $2(N/n - 1)$ generations ago.
Inverting this relationship between time and number of roots
implies that a marginal tree cut $T$ units of time ago
% since 2N(1/n - 1/N) = 2(N/n-1) = T implies
%    N/n = (T+2)/2 and so n = 2N/(T+2)
is expected to have around $r(T)$ roots, where $r(T) = 2N/(T+2)$.
The total tree length over this time is
$2N \sum_{k=r(T)+1}^N 1/(k-1) \approx 2 N \log(N/r(T))$.
This is a crude estimate for several reasons:
first, we should not count the branch leading to the root of the tree (i.e., when $r(T)=1$),
and second, this does not account for the discrete nature of the Wright--Fisher model.
% which is approximately
% \begin{align*}
%     2N - 2 + 8 N \left\{ \log(N-1) - \log(r(T)) \}
%        = 2N - 2 + 8 N \log( (T+2)(N-1)/2N )
%        \approx 2 N (1 + 4 \log( (T+2)/2 ))
% \end{align*}
Nonetheless, this leads as above to an upper bound on the number of edges of
\begin{align}
    \label{eqn:edge_bound}
    2 N \left( 1 + 4 \log\left( \min\left(N, \frac{T+2}{2}\right) \right)\right) .
\end{align}
This implies that the number of edges required to store the last $T$ generations
of history for the entire chromosome of a population of size $N$
grows as $O(N \log T)$ -- proportionally to $N$ at first but rapidly tapering off.
The bound holds up reasonably well in practice,
as shown in Figure~\ref{fig:simplify_complexity}.

@@

%%%%%%%%%%%%%%%%%%%%%%
\section*{Acknowledgements}
Thanks to Jared Galloway, Brad Shaffer, and Evan McCartney--Melstad for useful discussions.
Work on this project was supported by funding from
the Sloan Foundation and the NSF (under DBI-1262645) to PR;
the Wellcome Trust (grant 100956/Z/13/Z to Gil McVean);
the NIH (R01GM115564) to KRT.

>>>>

%% Progressive memory reduction as a function of T
What if $T$ is smaller:
how many of the resulting $2NT$ edges are required after simplification?
In other words, how fast does the information in the pedigree become irrelevant?
% \citet{padhukasahasram2008exploring} obtained impressive speedups relative to contemporary simulators
% by retaining only 8 generations of history.
% This begs the question: how fast does the information in the pedigree become irrelevant?
Now, we need to compute the expected total length of all branches in a coalescent tree
up until time $T$ (or the common ancestor, whichever comes first).
Again, coalescent theory tells us that
the expected length of time for which a coalescent tree has $k$ lineages
is $2N/(k(k-1)) = 2N(1/(k-1) - 1/k)$ generations.
Since the tree has $k$ branches over this period,
it is expected to contribute $2N/(k-1)$ to the total tree length.
By summing over $n < k \le N$,
the $N$ tips of a tree are expected to descend from only $n$ lineages
around $2(N/n - 1)$ generations ago.
Inverting this relationship between time and number of roots
implies that a marginal tree cut $T$ units of time ago
% since 2N(1/n - 1/N) = 2(N/n-1) = T implies
%    N/n = (T+2)/2 and so n = 2N/(T+2)
is expected to have around $r(T)$ roots, where $r(T) = 2N/(T+2)$.
The total tree length over this time is
$2N \sum_{k=r(T)+1}^N 1/(k-1) \approx 2 N \log(N/r(T))$,
since $\sum_{k=1}^x 1/k \approx \log(x) + \gamma$.
This is a crude estimate for several reasons:
first, we should not count the branch leading to the root of the tree (i.e., when $r(T)=1$),
and second, this does not account for the discrete nature of the Wright--Fisher model.
% which is approximately
% \begin{align*}
%     2N - 2 + 8 N \left\{ \log(N-1) - \log(r(T)) \}
%        = 2N - 2 + 8 N \log( (T+2)(N-1)/2N )
%        \approx 2 N (1 + 4 \log( (T+2)/2 ))
% \end{align*}
Nonetheless, this leads as above to an upper bound on the number of edges of
\begin{align}
    \label{eqn:edge_bound}
    2 N \left( 1 + 4 \log\left( \min\left(N, \frac{T+2}{2}\right) \right)\right) .
\end{align}
This implies that the number of edges required to store the last $T$ generations
of history for the entire chromosome of a population of size $N$
grows as $O(N \log T)$ -- proportionally to $N$ at first but rapidly tapering off.
The bound holds up reasonably well in practice,
as shown in Figure~\ref{fig:simplify_complexity}.

@@

%%%%%%%%%%%%%%%%%%%%%%
\section*{Acknowledgements}
Thanks to Jared Galloway, Brad Shaffer, and Evan McCartney--Melstad for useful discussions.
Work on this project was supported by funding from
the Sloan Foundation and the NSF (under DBI-1262645) to PR;
the Wellcome Trust (grant 100956/Z/13/Z) to Gil McVean;
the NIH (R01GM115564) to KRT.
